Manuscript_draft

### 2021-02-20 manuscript_draft.Rmd
working on manuscriot

# plot multiplier
pm=25

@@

### Rate of change from 1992 to 2019
All treatment units increased volume, basal area and stand density over the 1992-2019 period (`r figNums("standPlots",display="cite")` and `r appNums("standPlot.table",display="cite")`).  Mean annual basal area and volume increment peaked during the 1997-2009 period in the two harvested treatments, and peaked in the 1994-1997 period in the control treatment unit (`r figNums("perInc.plot",display="cite")`).  From 1992 to 2019, estimated marginal means of basal area increment in the low RBA unit exceeded marginal means in the high RBA and control unit by an estimated 
`r standFit.contrasts %>% as.data.frame() %>% filter(Variable=="Basal area") %>%   pull(estimate) %>% round(.,1) %>% .[1]`
and
`r standFit.contrasts %>% as.data.frame() %>% filter(Variable=="Basal area") %>%   pull(estimate) %>% round(.,1) %>% .[2]`
m² per hectare, respectively (p<0.01 for both, `r appNums("standFit.contrasts",display="cite")`).

@@

## Species differences in growth response to partial harvest

At the tree-level, our results suggest that while both hybrid white sruce and subalpine fir respond to thinning with faster radial growth, heavy thinning can elicit a stronger radial growth response in spruce, compared to fir.  This coincides with our finding of proportionally higher spruce volume and basal area in the heavy thinning treatment in 2019, compared to the other treatments.  Where thinning treatments in this region retain both species, removing a relatively high amount of the stand may favour spruce release.  Other studies have shown that residual spruce growth responds favourably to heavy thinning.  A 10-year analysis of white spruce growth underneath an aspen canopy found that a complete or high overstory removal resulted in immediate and significant spruce radial and height growth responses, compared to control and partial cutting treatments (Smith et al. 2016).   In a 35-year old white spruce stand, diameter increment was maximized with the heaviest thinning intensity, however, the treatment thinning from below (Stiell 1970).  Frank (1973) also showed that 70-year old white spruce basal increment increased with the amount of thinning. Competition for light is a key factor that limits white spruce growth, and opening up the stand to allow for increased light penetration may be a primary mechanism to enhance white spruce growth responses [@Lieffers1999].  High basal area removal may also increase stand air and soil temperatures, which can preferentially favour spruce growth responses compared to fir [@Lajzerowicz2004].

Other comparative studies of spruce and fir growth rates present differing evidence regarding species differences in release potential to thinning.  In mature Engelmann spruce (Picea engelmannii) – subalpine fir forests in western North America, understory spruce growth rates exceeded those of fir, allowing spruce to achieve canopy position faster, including after canopy gaps were created through disturbance (Antos et al. 2000, Andrus et al. 2018).  In mature forest in central BC, spruce growth exceeded that of subalpine fir after age 80 (Eis and Craigdallie 1983).  Because spruce tends to live longer than subalpine fir, higher growth rates at later ages may allow this species to become larger than fir.  In contrast, thinning studies in spruce-fir forests in eastern North America report that balsam fir (*Abies balsamifera*) responds to release with faster height and diameter growth than co-occurring red spruce (Picea rubens) (Dumais and Prevost, 2007; Dumais et al. 2014, Messier et al. 1999). Although both species are similar in shade tolerance, firs are considered to be better adapted to low light conditions in the understory, allowing them to persist longer in undisturbed forest  (Antos 2000) and potentially able to respond faster to changes in understory light (Dumais et al 2014, Messier et al. 1999).  Other studies have found no difference in growth response between spruce and fir species after treatment (MCaughey and Smith1982).   Our study also suggests that at lower levels of basal area removal, spruce and fir growth responses will be similar.

## Management considerations
1. Both species responded with increased growth after partial harvest.  High removal of basal area in these forests stimulated stronger spruce growth.  This information can help silviculturists manipulate stand structure to preferentially ...
2. Partial harvesting at both intensities delayed peak volume and basal area increment by approximately 10 years.  Annual increment in the control unit declined faster over the last measurement period compared to previous periods, perhaps suggesting that the stand is becoming mature and losses due to mortality may substantially offset growth increment.

>>>>

# plot multiplier
pm=20

@@

### Rate of change from 1992 to 2019
All treatment units, including the control, increased volume, basal area and stand density over the 1992-2019 period (`r figNums("standPlots",display="cite")` and `r appNums("standPlot.table",display="cite")`).  Mean annual basal area and volume increment peaked during the 1997-2009 period in the two harvested treatments, and peaked in the 1994-1997 period in the control treatment unit (`r figNums("perInc.plot",display="cite")`).  From 1992 to 2019, estimated marginal means of basal area increment in the low RBA unit exceeded marginal means in the high RBA and control unit by an estimated 
`r standFit.contrasts %>% as.data.frame() %>% filter(Variable=="Basal area") %>%   pull(estimate) %>% round(.,1) %>% .[1]`
and
`r standFit.contrasts %>% as.data.frame() %>% filter(Variable=="Basal area") %>%   pull(estimate) %>% round(.,1) %>% .[2]`
m² per hectare, respectively (p<0.01 for both, `r appNums("standFit.contrasts",display="cite")`).

@@

## Tree-level growth response to partial harvest

At the tree-level, our results suggest that while both hybrid white sruce and subalpine fir respond to thinning with faster radial growth, heavy thinning can elicit a stronger radial growth response in spruce, compared to fir.  This coincides with our finding of proportionally higher spruce volume and basal area in the heavy thinning treatment in 2019, compared to the other treatments.  Where thinning treatments in this region retain both species, removing a relatively high amount of the stand may favour spruce release.  Other studies have shown that residual spruce growth responds favourably to heavy thinning.  A 10-year analysis of white spruce growth underneath an aspen canopy found that a complete or high overstory removal resulted in immediate and significant spruce radial and height growth responses, compared to control and partial cutting treatments (Smith et al. 2016).   In a 35-year old white spruce stand, diameter increment was maximized with the heaviest thinning intensity, however, the treatment thinning from below (Stiell 1970).  Frank (1973) also showed that 70-year old white spruce basal increment increased with the amount of thinning. Competition for light is a key factor that limits white spruce growth, and opening up the stand to allow for increased light penetration may be a primary mechanism that enhances white spruce growth responses [@Lieffers1999].  High basal area removal may also increase air and soil temperatures, which can preferentially favour spruce growth responses compared to fir [@Lajzerowicz2004].

Other comparative studies of spruce and fir growth rates present differing evidence regarding species differences in release potential to thinning.  In mature Engelmann spruce (*Picea engelmannii*) – subalpine fir forests in western North America, understory spruce growth rates exceeded those of fir, allowing spruce to achieve canopy position faster, including after canopy gaps were created through disturbance (Antos et al. 2000, Andrus et al. 2018).  In mature forest in central BC, spruce growth exceeded that of subalpine fir after age 80 (Eis and Craigdallie 1983).  Because spruce tends to live longer than subalpine fir, higher growth rates at later ages may allow this species to become larger than fir.  In contrast, thinning studies in spruce-fir forests in eastern North America report that balsam fir (*Abies balsamifera*) responds to release with faster height and diameter growth than co-occurring red spruce (*Picea rubens*) (Dumais and Prevost, 2007; Dumais et al. 2014, Messier et al. 1999). Although both species are similar in shade tolerance, firs are considered to be better adapted to low light conditions in the understory, allowing them to persist longer in undisturbed forest  (Antos 2000) and potentially able to respond faster to changes in understory light (Dumais et al 2014, Messier et al. 1999).  Other studies have found no difference in growth response between spruce and fir species after treatment (MCaughey and Smith1982).   Our study also suggests that at lower levels of basal area removal, spruce and fir growth responses are similar.

In both species, post-treatment growth rates increased with the tree size at the time of treatment.  This has been found in other studies as well. and is likely explained by the assymetric competition theory. Implications of this result are that growth increased from treatment will be concentrated in the larger tree size classes.  If larger trees are able to increase their size faster than smaller trees and increase other metrics, such as crown size, to take advantage of increased growing space, then smaller trees will continue to experience competition and will not benefit as much from the thinning treatment.  Forest managers should account for this size differential when planning thinning treatments.  For example, a silviculture treatment that seeks to release a suppressed understorey may have to ensure that sufficient overstorey trees are removed to reduce assymetric competition after treatment.

## Stand-level responses to thinning treatments

Volume

Diameter distributions
All units in this stand maintained a negative-exponential diameter distribution whereby the number of trees was highest in the smallest diameter classes.  Insofar as this diameter distribution indicates an uneven-aged stand structure, thinning in these stands maintained these characteristics and did not substantially alter the structure.  The species composition between the low removal and control units was also similar, in that subalpine fir was the leading species in the smaller diameter classes, and spruce was leading the species in the larger classes.  This may reflect species differences in regeneration and growth.  In closed forests with undisturbed forest floors, Subalpine fir can have higher seedling establishment success than spruce, leading to higher densities of fir in the understorey.  Subalpine fir can be relatively short-lived, however, and can die after 250 years in the canopy.  Spruce, on the other hand, can live for over 400 years and thus can achieve larger sizes and dominant canopy positions than fir.  This leads to a common stand structure in the Interior Region of BC that was found in the low-reomval and control untis.

The high-removal unit had a somewhat different species composition.  Here, fir was the leading sprecies in all diameter classes.  This means that the high-removal thinning treatment converted the stand compsition to fir-leading, which is a primary concern for many foresters in British Columbia.  Spruce is generally much preferred as a crop species over subalpine fir.  However, the stands in this treatment had more spruce in the smaller diameter classes than the other treatment units and our results show that spruce increased its composition compared to fir over the 27-year period of this study.  In contrast, the ratio of spruce to fir volume decreased over the same period in both the low-removal and control units.  Over a timespan of several decades, it seems reasonable to anticipate that spruce:fir ratios will converge between all treatment units.

Stand mortality increased with basal area after treatment.  This suggests that the thinning treatments may have captured some volume through trees that would have died naturally.  Thus, while volume removals from thinning are never entirely recoverable, this may not be true when mortality is considered.

## Management considerations
1. Both species responded with increased growth after partial harvest.  High removal of basal area in these forests stimulated stronger spruce growth.  This information can help silviculturists manipulate stand structure to preferentially ...
2. Partial harvesting at both intensities delayed peak volume and basal area increment by approximately 10 years.  Annual increment in the control unit declined faster over the last measurement period compared to previous periods, perhaps suggesting that the stand is becoming mature and losses due to mortality may substantially offset growth increment.
3. GIven that larger trees responded with faster growth than smaller trees, silviculturists need to consider the implications of this on residual stand structure and response after treatment.
